Stagnating State Funding for Higher Education and its Effect on Faculty Composition at US Universities

### 2023-05-30 text/sections/00-abstract.tex
Finished round of writing revisions.  Only spell-checking remains.

While states have limited their funding for higher education for the last forty years, US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers.
This analysis relates the changes the faculty composition at US public universities to the fall in state funding for higher education, using a shift-share approach to identify causal effects.
A $-10$\% fall in a university's funding, via a state funding shock, decreases the number of assistant professors per student by $-1.4$\% and full professors by $-1.2$\%, yet increases the number of lecturers per student by 4.4\%; long-run estimates show the effects linger for multiple years after the initial funding shock.
Analysis of all the professors at Illinois public universities shows that incumbent professors are not affected by substantial funding shocks over 2011-2021, and suggestive evidence shows that the faculty changes arose by falling hires of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

>>>>

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while at the same time states have limited how much they fund higher education.
This analysis relates the changes the faculty composition at US public universities to the fall in state funding for higher education, using a shift-share approach to identify causal effects.
A fall of $-10$\% in university funding, via a state funding shock, decreases the number of assistant professors per student by $-1.4$\% and full professors by $-1.2$\%, yet increases the number of lecturers per student by 4.4\%; long-run estimates show the effects linger for multiple years after the initial funding shock.
Analysis of all the professors at Illinois public universities shows that incumbent professors are not affected by substantial funding shocks over 2011-2021, with suggestive evidence showing that the faculty changes arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

### 2023-05-30 text/sections/01-introduction.tex
Finished round of writing revisions.  Only spell-checking remains.

Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last three decades.
At the same time, public university student enrollment and employment of contingenet lecturers dramtically increased at public universities.
I show that four-year, degree-granting public universities have experienced a systematic fall in state funding per student, and this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Falls in state funding for public universities, instrumented by state-level shocks to state-wide higher education, led to a fall in the number of tenure-track/tenured professors per student within a university, and an increase in the number of lecturers.
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state support for public universities has implications for the wider structure of US higher education and research.

US universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average public university received around \$11,600 of state funding per student in 1990, and only \$8,300 in 2021.
This fall is driven by a stagnation in the absolute level of state funding for higher education in most states, combined with a large increase of 46\% in student enrolment at public universities over the same time period.
Similarly, the number of lecturers per student over doubled, while the number of tenure-track/tenured professors per students fell by 23\% from 1990 to 2021.

I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education, and everything else, by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, it is necessary to employ an identification strategy, such as the shift-share instrument, to account for endogeneity in universities' state funding.
The shift-share instrument interacts reliance on state funding in a base period with the entire state' total funding per student for each year, to exploit how much each university historically relied on state funding and yearly changes in state funding for higher education.

Falls in state funding affects faculty composition at public universities, who substitute away from tenured and tenure-track professors towards contingent lecturers.
A fall of $-10$\% in state funding leads to a fall of $-1.4$\% in the number of assistant professors per student at a university, a fall of $-1.2$\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the immediate year of the first funding shock. 
Over the same time period, state funding fell by around $-$35\%, while the count of professors per student fell by $-$9\%, so that these results show that falls in state funding explain around a third of the substitution away from tenure-track and tenure professors towards contingent lecturers.
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet, the hiring rate for professors declined over this time period for all public universities, and particularly for universities with the greatest falls in state funding.
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

Absent this analysis, it is not immediately clear why stagnating state support and changes in public university faculty composition would be causally linked.
Over the 21st Century there have been drastic changes to college selectivity, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, so that over time public universities (on average) substitute towards contingent lecturers. 
This is one way in which the trends may be concurrent, but not causal.
This analysis shows that stagnating in state funding for public universities cause the changes in faculty composition, explaining about 40\% of the observed substitution towards contingent lecturers, so that it is possible that other systematic changes in higher education (such as changing selectivity) may explain the rest of the variation.

@@

Faculty composition is an interesting outcome on its own, but its relationship with higher education finances has so far not been studied.
The closest previous examples study  how university endowments react to negative endowment shocks \citep{brown2014endowment} and the 2008 recession \citep{turner2014impact}, and multiple theoretically model university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
%Lastly, instruction costs within individual universities vary substantially across subject areas, primarily thanks to department differences in class size and faculty salaries \citep{hemelt2021math}.
The number of professors per student may be particularly important for quality of instruction and thus educational outcomes.
Small class sizes in secondary schools lead to increases in test scores \citep{angrist1999using}, and UK university students perform worse academically in particularly large classes \citep{bandiera2010heterogeneous}.
Though, the economics literature is not decided on whether contingent lecturers lead to better or worse outcomes in higher education.
Adjunct lecturers are generally more effective teachers than research professors at highly selective universities \citep{bettinger2010does,figlio2015tenure}, but students taught by part-time lecturers have worse student outcomes relative to full-time lecturers or faculty \citep{zhu2021limited,ehrenberg2005tenured}.

While it is not immediately clear whether a systematic substitution towards lecturers away from professors reduces teaching effectiveness, it is clear that stagnating state funding has negatively impacted student outcomes \citep{NBERw23736,NBERw27885}.
This analysis focuses on a university's count of faculty per student, and considers faculty composition change as one possible mechanism for how stagnating state funding for higher education has negatively affected students at public universities for the last 30 years.

>>>>

Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) over the last three decades.
At the same time, both public university student enrolment and employment of contingent lecturers dramatically increased at public universities.
I show that four-year, degree-granting public universities have experienced a systematic fall in state funding per student, and this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Falls in state funding for public universities, instrumented by state-level shocks to state-wide higher education, led to a fall in the number of tenure-track/tenured professors per student within a university, and an increase in the number of lecturers.
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state support for public universities has implications for the wider structure of US higher education and research.

US universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average public university received around \$11,600 of state funding per student in 1990, and only \$8,300 in 2021.
This fall is driven by a stagnation in the absolute level of state funding for higher education in most states, combined with a large increase of 46\% in student enrolment at public universities over the same time period.
Similarly, the number of lecturers per student over doubled, while the number of tenure-track/tenured professors per students fell by $-23$\% from 1990 to 2021.

I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education, and everything else, by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, it is necessary to employ an identification strategy, such as the shift-share instrument, to account for endogeneity in universities' state funding.
The shift-share instrument interacts reliance on state funding in a base period with the entire state's total funding per higher education student for each year, to exploit changes in funding and how much each university historically relied on state funding for higher education.

Falls in state funding affects faculty composition at public universities, which substitute away from tenured and tenure-track professors towards contingent lecturers.
A fall of $-10$\% in state funding leads to a fall of $-1.4$\% in the number of assistant professors per student at a university, a fall of $-1.2$\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the immediate year of the first funding shock. 
Over the same time period, state funding fell by around $-$35\%, while the count of professors per student fell by $-$9\%, so that these results show that falls in state funding explain around a third of the substitution away from tenure-track and tenure professors towards contingent lecturers.
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at publicuniversities was negatively impacted by the falls in state funding (and by the funding shocks).
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

@@

Over the last half century, there have been drastic changes to the heirarchy of US higher education, where the top universities have become more selective for student applicants, and the average university less selective \citep{hoxby2009changing}.
Similarly, research faculty at US universities are increasingly concentrated among the graduates of highly selective PhD programmes \citep{wapman2022quantifying}.
Faculty composition is an interesting outcome on its own, but its relationship with higher education finances has so far not been studied.
The closest previous examples study  how university endowments react to negative endowment shocks \citep{brown2014endowment} and the 2008 recession \citep{turner2014impact}, and multiple theoretically model university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
%Lastly, instruction costs within individual universities vary substantially across subject areas, primarily thanks to department differences in class size and faculty salaries \citep{hemelt2021math}.
The number of professors per student may be particularly important for quality of instruction and thus educational outcomes;
small class sizes in secondary schools lead to increases in test scores \citep{angrist1999using}, and UK university students perform worse academically in particularly large classes \citep{bandiera2010heterogeneous}.
Though, the economics literature is not decided on whether contingent lecturers lead to better or worse outcomes in higher education.
Adjunct lecturers are generally more effective teachers than research professors at highly selective universities \citep{bettinger2010does,figlio2015tenure}, but students taught by part-time lecturers have worse student outcomes relative to full-time lecturers or faculty \citep{zhu2021limited,ehrenberg2005tenured}.

While it is not immediately clear whether a systematic substitution towards lecturers away from professors reduces teaching effectiveness, it is clear that stagnating state funding has negatively impacted student outcomes \citep{NBERw23736,NBERw27885}.
This analysis focuses on a university's count of faculty per student, and considers faculty composition change as one of the mechanisms for how stagnating state funding for higher education transmits to students and has led to worsening education outcomes at public universities for the last thirty years.

### 2023-05-30 text/sections/02-data.tex
Finished round of writing revisions.  Only spell-checking remains.

The decade of 2010--2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015 partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, so is the same as the academic year definition.
}
While public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This variation in public university revenues via the state funding channel, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- mean that Illinois public universities exhibit sizable changes in their state funding over 2010-2021, of similar order to those for the rest of the country over 1990--2021.

>>>>

The decade of 2010--2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015, partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, so is the same as the academic year definition.
}
While public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This variation in public university revenues via the state funding channel, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- mean that Illinois public universities exhibit sizable changes in their state funding over 2010-2021, of similar order to those for the rest of the country over 1990--2021.

### 2023-05-30 text/sections/03-concepts.tex
Finished major revision.  Needs more checks of writing style.
Adjusting writing in TeX document.  2nd revision unfinished.
Second major revision: Concepts needs overhaul.
Finished round of writing revisions.  Only spell-checking remains.

Most universities hire a number of new professors every year, either to expand their departments or to replace leaving professors.
The number of professors hired in each year is usually highest among non-tenured adjunct faculty, as these instructors are mostly hired on short-term contingent contracts, so that they are less costly to hire (or to not reinstate) in response to yearly changes in teaching needs.
Secondly, tenure-track assistant professors are hired to replace leaving assistant professors, and are most often hired with a four to six year contract and agreement for tenure consideration.
Lastly, tenured professors have undergone review and have successfully secured a full-time appointment at their university with no expiration date; tenured professors may be hired from outside a university, though this position has the lowest yearly hiring rate among most universities.
It is common for universities to restrict their faculty hiring in the face of financial shocks.
Since the yearly hiring rate is highest among lecturers and assistant professors, financial shocks may lower the number of lecturer and assistant professors at public universities the most, if faculty hiring is affected by falls in funding.

Additionally, faculty salaries may by affected by changes in public university finances.
When the university has a lower budget for its yearly hiring, it may respond by lowering the salary they offer to new hires.
This possible effect may not be the same across each position of professor; tenured faculty are often hired away from another university, so that new tenured professor hires may be less likely to accept a lower salary offer from a public university.\footnote{
    Faculty often consider their outside options, as in offers for employment and/or promotion from other universities.
    See \cite{blackaby2005} for an overview of faculty outside options.
}
Yet salary for all the professors, not just new hires, may also be affected: multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, for example.\footnote{
    Indeed, Cornell University implemented hiring freezes and a nominal salary reduction for professors, in anticipation of a financial shock in early 2021.
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, so that the cut was returned to professors in the year 2021.
}
Faculty are not paid the same by position, so that there are financial consequences for a university promoting their own faculty.
In response to tightening fiscal restraints, universities may be less likely to grant tenure to assistant professors (as tenured professors generally receive a salary raise and long term contract), or to promote tenured professors from associate to full professor.

It is not immediately clear which effects will dominate, or whether tenured professors vs lecturers will be affected more by changes in their university's funding.
Yet, there is one empirical fact worth noting: a mean assistant or tenured professor earns more than double that of an adjunct lecturer per year (see \autoref{tab:illinois-summary}), and the tenure contract means universities must pay a higher salary for a guaranteed number years, more so than for contingent lecturers.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer and fewer resources, then they may substitute away from tenure-track and tenured professor towards contingent lecturers to achieve this obligation.\footnote{
    That is, substituted away from tenure-track and tenured professors according to the salary ratio, and the relative productivity between these two positions.
}

>>>>

Universities hire new professors most years, either to expand their departments or to replace leaving professors.
The number of professors hired in each year is usually highest among non-tenured adjunct faculty, as these instructors are mostly hired on short-term contingent contracts, so that they are less costly to hire (or to not reinstate) in response to yearly changes in teaching needs.
Tenure-track assistant professors are hired most often hired with a four to six year contract, and formal agreement for tenure consideration at the end of this term.
Lastly, tenured professors have undergone review and have successfully secured a full-time appointment at their university with no expiration date; tenured professors may be hired from outside a university, though this position has the lowest yearly hiring rate among most universities.
It is common for universities to restrict or delay their faculty hiring in the spring term ahead of the next academic year, in response to unforeseen events, or changes in needs.
So it stands to reason that the hiring rate of new professors could be meaningfully affected by changes in university finances, and thus falls in state funding.

%Additionally, faculty salaries may be affected by changes in public university finances.
%When the university has a lower budget for its yearly hiring, it may respond by lowering the salary they offer to new hires.
%This possible effect may not be the same across each position of professor; tenured faculty are often hired away from another university, so that new tenured professor hires may be less likely to accept a lower salary offer from a public university.\footnote{
%    Faculty often consider their outside options, as in offers for employment and/or promotion from other universities.
%    See \cite{blackaby2005} for an overview of faculty outside options.
%}
%Yet salary for all the professors, not just new hires, may also be affected: multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, for example.\footnote{
%    Indeed, Cornell University implemented hiring freezes and a nominal salary reduction for professors, in anticipation of a financial shock in early 2021.
%    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, so that the cut was returned to professors in the year 2021.
%}
%Faculty are not paid the same by position, so that there are financial consequences for a university promoting their own faculty.
%In response to tightening fiscal restraints, universities may be less likely to grant tenure to assistant professors (as tenured professors generally receive a salary raise and long term contract), or to promote tenured professors from associate to full professor.

Absent this analysis, it is not immediately clear why stagnating state support and changes in faculty composition at public universities would be causally linked.
Over the 21st Century there have been drastic changes to college selectivity, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, so that over time public universities (on average) substitute towards contingent lecturers. 
This is one way in which the trends may be concurrent, but not causal.
This analysis shows that stagnating in state funding for public universities cause the changes in faculty composition, explaining about 40\% of the observed substitution towards contingent lecturers, so that it is possible that other systematic changes in higher education (such as changing selectivity) may explain the rest of the variation.

It is not immediately clear which effects will dominate, or whether tenured professors or lecturers will be affected more by changes in their university's funding.
Yet there is one empirical fact worth noting: the average assistant or tenured professor earns more than double that of a lecturer (see \autoref{tab:illinois-summary}).
Additionally, universities must pay the higher salary to tenure-track/tenured professors for a guaranteed number years, which is not true for lecturers who are mostly on shorter term contracts.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer and fewer resources, then they may substitute away from tenure-track and tenured professor towards contingent lecturers.
